---
title: "The Transformer and the Road to Modern LLMs"
subtitle: "Gentle Introduction to AI (HVEE.00.049) -- Week 06"
author: 
    name: Maciej Eder
    affiliation: University of Tartu | Polish Academy of Sciences
date: 2026/09/08
format: 
  revealjs:
    theme: pp.scss
    logo: DigiTS-RGB-short-SMALL.svg
    css: big_logo_DigiTS.css
    self_contained: false
---

# Part I. Why Transformers Were Needed


## Recap: from Week 5

| Model | Main idea |
|-------|-- -----|-----|
| word2vec | word meaning from context |
| BERT | contextual embeddings |
| **Transformer** | **attention-based architecture** |
| GPT | Transformer scaled for generation |


## The problem with RNNs / LSTMs

- They process text **sequentially**, one word at a time
- Long-range dependencies are hard to learn
- Gradient vanishing: distant words "forget" what came before

### Key intuition

> The book that I bought _yesterday_ because my old copy was damaged was excellent.

To understand *book* ↔ *excellent*, the RNN must propagate info through many intermediate words.


## The bottleneck

| Architecture | Parallel processing? | Long-range context? |
|----|------|--------------------|
| RNN/LSTM | No | Poor |
| Transformer | **Yes** | **Excellent** |


# Part II. Attention as Selective Reading


## The attention question

When humans read:

> The king spoke to his **subjects**.

which words matter for interpreting "king"?

Students naturally answer:

* king
* subjects
* spoke

Attention = selective focus on relevant context.


## A DH analogy

Attention resembles:

* **citation networks** — not all references are equally important
* **intertextual references** — some texts echo earlier works more than others
* **close reading** — certain words carry more interpretive weight

> Not all words contribute equally to meaning.


# Part III. Self-Attention


## How self-attention works

A token asks:

> Which other tokens in this sentence should I pay attention to?

Example:

> The animal didn't cross the street because **it** was too tired.

* **it** = ?
* The animal ← this is what matters for *it*


## Key insights

Attention allows:

* **long-range dependencies** — far-apart words still connect
* **parallel computation** — no need to wait for sequential processing
* **context-sensitive representations** — same token, different focus


## Visualising attention

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle, fixedsize = true, width = 2, style = filled, fillcolor = Beige]
  node [fontsize = 13]
  
  A1 [label = 'The']
  A2 [label = 'king']
  A3 [label = 'spoke']
  A4 [label = 'to']
  A5 [label = 'his']
  A6 [label = 'subjects']
  
  B1 [label = 'A1']
  B2 [label = 'A2']
  B3 [label = 'A3']
  B4 [label = 'A4']
  B5 [label = 'A5']
  B6 [label = 'A6']
  
  edgesetA [label = 'A', shape = plaintext]
  
  subgraph cluster_attention {
    label = 'Attention Weights'
    style = rounded
    A1 -> A2 [color=blue, penwidth=3, label='strong']
    A1 -> A3 [color=blue, penwidth=2]
    A1 -> A6 [color=blue, penwidth=1]
    A1 -> A4 [color=grey, penwidth=0.5, style=dotted]
    A1 -> A5 [color=grey, penwidth=0.5, style=dotted]
    A6 -> A2 [color=red, penwidth=4, label='focus']
    A6 -> A3 [color=red, penwidth=2]
    A6 -> A1 [color=red, penwidth=1]
  }
  
  }")

```

# Part IV. Multi-Head Attention


## Multiple readers

Imagine several readers analyzing the **same text** simultaneously:

1. **Reader 1** focuses on *syntax* (grammar roles)
2. **Reader 2** focuses on *semantic relations* (meaning)
3. **Reader 3** focuses on *discourse structure* (how ideas connect)
4. **Reader 4** focuses on *named entities* (people, places)

Each is an **attention head**.


## One sentence, several perspectives

> The old king who ruled the northern provinces spoke to the ambassador.

- Head 1 finds: **king** → **spoke** (subject-verb)
- Head 2 finds: **provinces** ↔ **northern** (modifier-noun)
- Head 3 finds: **king** ↔ **oldest** (age attribute)
- Head 4 finds: **ambassador** as a person


## Visual comparison

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = TD]
  
  node [shape = rectangle, style = filled, width = 3.5]
  
  box1 [label = 'input\\ntokens', fillcolor = Beige, width = 4]
  
  node [shape = rectangle, style = filled, width = 2]
  
  h1 [label = 'head 1\\nsyntax', fillcolor = LightBlue]
  h2 [label = 'head 2\\nsemantics', fillcolor = LightGreen]
  h3 [label = 'head 3\\ndiscourse', fillcolor = LightYellow]
  h4 [label = 'head 4\\nentities', fillcolor = MistyRose]
  
  merged [label = 'concatenate\\n& project', fillcolor = Linen, width = 4]
  
  box1 -> h1
  box1 -> h2
  box1 -> h3
  box1 -> h4
  
  h1 -> merged
  h2 -> merged
  h3 -> merged
  h4 -> merged
  
  }")

```

# Part V. Positional Encoding


## An obvious problem

> Transformers process words **simultaneously**.
> How do they know the **order**?


## The dog bites man / man bites dog test

| Sentence | Meaning |
|------------------------|------------------------------|
| Dog bites man | The dog is the aggressor |
| Man bites dog | The man is the aggressor |

Same tokens, **different meaning**.


## Positional encoding

* Injects **order information** into token representations
* No mathematics needed for intuition:
  each token gets a unique "timestamp" added to its vector

> Attention knows **relationships**; positional encoding knows **sequence**.

# Part VI. Encoder vs Decoder


## The architecture landscape

| Model | Architecture | Use |
|-------|-- -----|-----|
| BERT | **Encoder** | Understanding (classification, NER...) |
| T5 | Encoder-Decoder | Translation, summarization |
| **GPT** | **Decoder-only** | Text generation |
| Mistral | Decoder-only | Text generation |
| DeepSeek | Decoder-only | Text generation + reasoning |


## Why different architectures?

| Architecture | Strength |
|----|-------|
| **Encoder** | Reads full context → excellent for **understanding** |
| **Decoder** | Generates one token at a time → excellent for **creating** |
| **Encoder-Decoder** | Reads full input → generates structured output → excellent for **translation** |


## BERT vs GPT

| Feature | BERT | GPT |
|---------|------|-----|
| Architecture | Encoder | Decoder |
| Direction | Bidirectional | Unidirectional (left to right) |
| Training | Masked language modeling | Next-token prediction |
| Primary use | Classification, understanding | Generation, chatbots |
| DH use | Semantic search, NER, text analysis | Text generation, summarization |

# Part VII. GPT in Practice


## GPT's core idea

> The capital of Estonia is ...

1. Predict next token
2. Append it to the text
3. Repeat

Do this **thousands of times**.

Scale dramatically.

The result is surprisingly sophisticated behavior.


## The decoder pipeline

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle, style = filled, fillcolor = Beige]
  
  box1 [label = 'input\\nprompt']
  box2 [label = 'DECODER\\nLAYERS', fillcolor = Linen, width = 3]
  box3 [label = 'next token\\nprobability']
  box4 [label = 'sample / argmax']
  box5 [label = 'new token']
  
  # edges
  box1 -> box2
  box2 -> box3
  box3 -> box4
  box4 -> box5
  box5 -> box2
  
  # make it look like a loop
  box5 -> box4 [dir=back, style=dotted, color=grey]
  box5 -> box1 [label = 'appended', style = dashed]
  
  }")

```

# Part VIII. Scaling Laws


## Three dimensions of scale

| Dimension | What it is |
|----|------|
| **Model size** | Number of parameters |
| **Data size** | Amount of training text |
| **Compute** | Training resources (GPU time) |


## The Kaplan result (2020)

> Performance improves **predictably** with scale.
> No obvious plateau was observed.

This was one of the discoveries that triggered the current LLM boom.

**Surprise**: bigger isn't worse -- bigger is **more capable**.


## DH analogy

Consider:

* A corpus of **1,000 books**
* A corpus of **10 million books**

The larger corpus reveals patterns unavailable in the smaller one.

**Larger data → richer representations → emergent capabilities.**


## Emergent abilities

As models scale, new capabilities appear **not designed** but **emergent**:

| Capability | Appears around |
|--|--|
| In-context learning | GPT-3 scale |
| Chain-of-thought reasoning | GPT-4 scale |
| Code generation | GPT-4 scale |
| Few-shot learning | GPT-3 scale |


# Part IX. The Modern LLM Ecosystem


## GPT (OpenAI)

* Established the **decoder-only** paradigm for chat/generation
* GPT-4: multimodal, millions of parameters
* Proprietary, API-based

## Mistral (Mistral AI)

* Shows that **efficient** open-weight models can compete
* Mistral 7B -- competitive with much larger models
* Open weights → usable locally

## DeepSeek

* Demonstrates how **architectural optimization** reduces costs
* Training strategy focus
* Strong open-source reasoning models
* Efficient training at scale

## The key message

> Modern LLM development is **not** about entirely new architectures.
> It's about **efficiency, training strategy, and scaling**.


# Part X. Humanities Perspective


## What changes when archives are "understood"?

1. **Semantic search** replaces keyword matching
2. Every historical document becomes **queryable**
3. **Latent representations** replace manual classification
4. Patterns invisible to human readers become visible

## The DH opportunity

| Before | After Transformers |
|--------|----|
| Keyword search | Semantic search |
| Manual annotation | Model-assisted labeling |
| Small corpus research | Full-archive analysis |
| Close reading | **Close reading + distant reading together** |


# Summary


## Key takeaways

1. **Transformers solved a real problem** -- RNN limits were a bottleneck
2. **Attention is selective focus** -- not all words matter equally
3. **Multi-head attention = many perspectives** on the same text
4. **Positional encoding** restores order information
5. **Encoder = understanding, Decoder = generation**
6. **Scaling is the secret** -- more data, more parameters, more compute
7. **Modern LLMs share one architecture** -- the decoder-only Transformer


## Recommended Readings

### Essential

1. **The Illustrated Transformer**
   [Jay Alammar](https://jalammar.github.io/illustrated-transformer/)
   Assign the entire article.

2. **Attention Is All You Need** (Vaswani et al. 2017)
   [arXiv:1706.03762](https://arxiv.org/abs/1706.03762)
   Assign: Abstract, Introduction, Figure 1, Section 3.2

3. **The Illustrated GPT-2**
   [Jay Alammar](https://jalammar.github.io/illustrated-gpt2/)
   The essential bridge from Transformer to LLM.

### Strongly Recommended

4. **Language Models are Few-Shot Learners** (Brown et al. 2020)
   [arXiv:2005.14165](https://arxiv.org/abs/2005.14165) -- Abstract, Introduction, Scaling discussion

5. **Scaling Laws for Neural Language Models** (Kaplan et al. 2020)
   [arXiv:2001.08361](https://arxiv.org/abs/2001.08361) -- selected excerpts

### Optional

6. **The Annotated Transformer**
   [nlp.seas.harvard.edu/annotated-transformer](https://nlp.seas.harvard.edu/annotated-transformer/)

7. **Mistral 7B**
   [arXiv:2310.06825](https://arxiv.org/abs/2310.06825)

8. **DeepSeek-V3 Technical Report**
   [arXiv:2412.19437](https://arxiv.org/abs/2412.19437)

# Thank you!
