---
title: "Multivariate classification"
subtitle: "Gentle Introduction to AI (HVEE.00.049) -- Week 01"
author: 
    name: Maciej Eder
    affiliation: University of Tartu | Polish Academy of Sciences
date: 2026/09/22
format: 
  revealjs:
    theme: pp.scss
    logo: DigiTS-RGB-short-SMALL.svg
    css: big_logo_DigiTS.css
    self_contained: false
---



```{r setup, include=FALSE}
# set global chunk options
library(knitr)

my.blue = rgb(0.15, 0.45, 0.96)
my.green = rgb(0.15, 0.85, 0.27, 0.7)
my.red = rgb(0.92, 0.3, 0.3, 0.6)
my.grey = rgb(0,0,0,.6)
my.orange = rgb(1,0.5,0.2,0.6) 
my.teal = rgb(0, 0.5, 0.5, 0.7)  #### or: my.teal = rgb(0.24, 0.65, 0.75, 0.7)
my.violet = rgb(0.75, 0.25, 0.82, 0.7)

my.p.blue = "rgba(38, 115, 245, 0.7)"
my.p.green = "rgba(38, 217,  69, 0.7)"
my.p.red = "rgba(235, 77, 77, 0.6)"
my.p.grey = "rgba(0, 0, 0, 0.6)"
my.p.orange = "rgba(255, 128, 51, 0.6)"
my.p.teal = "rgba(0, 128, 128, 0.7)"
my.p.violet = "rgba(190, 64, 210, 0.7)"

opts_chunk$set(cache=FALSE)
```

# recap

## the concepts introduced so far

- variable (feature)
- model
- linear regression
- logistic regression
- prediction
- (binary) classification


## simple dataset

```{r}
set.seed(4321)
n <- 200
hours <- runif(n, 0, 20)
prob_pass <- 1 / (1 + exp(-(-4.8 + 0.65 * hours)))
passed <- rbinom(n, size = 1, prob = prob_pass)
df <- data.frame(Hours = round(hours, 1), Passed = passed)
df[1:50,]
```


## dataset plotted

```{r}
plot(df$Hours, df$Passed, ylab = "probability of passing", xlab = "effort")
```

## clearly two classes

```{r}
colors = df[,2]
colors[colors == 1] = my.green
colors[colors == 0] = my.red
plot(df$Hours, df$Passed, col = colors, ylab = "probability of passing", xlab = "effort")
```


## logistic regression model

```{r}
model <- glm(Passed ~ Hours, data = df, family = binomial(link = "logit"))
plot(df$Hours, df$Passed, col = colors, ylab = "probability of passing", xlab = "effort")
lines(sort(df$Hours), predict(model, newdata=data.frame(Hours = sort(df$Hours)), type = "response"), col = "blue", lwd = 2)
```


## the decision line

```{r}
model <- glm(Passed ~ Hours, data = df, family = binomial(link = "logit"))
plot(df$Hours, df$Passed, col = colors, ylab = "probability of passing", xlab = "effort")
lines(sort(df$Hours), predict(model, newdata=data.frame(Hours = sort(df$Hours)), type = "response"), col = "blue", lwd = 2)
x0 <- -coef(model)[1]/coef(model)[2]
points(x0, 0.5, pch = 20, cex = 3)
abline(v = x0, lty = 2)
text(9, 0.5, labels= round(x0, 4))
```


## binary classification

- spam --- no-spam
- fake news --- genuine fact
- pass --- fail
- tsunami --- no danger
- enemy (target) --- ally (don't shoot)
- human --- vampire
- will break --- won't break
- original Shakespeare --- false attribution



# authorship attribution



## non-traditional authorship attribution

Given:

* a text of uncertain or anonymous authorship, and
* a comparison corpus of texts by known authors,

is it possible to find the **'nearest neighbor'**
among the available candidates?


## authorial fingerprint

- an idea of stylistic fingerprint:
    - undiscoverable with the naked eye
    - beyond authorial control
    - resistant to imitation, plagiarism and parody
    - popular solution: frequencies of the Most Frequent Words
- is the style _determined_ by an individual? (cf. human DNA code, fingerprint, or patterns in one’s iris)


## example: _Prometheus Bound_

* Aeschylus known for his 7 tragedies
* _Prometheus Bound_ somewhat odd, and thus disputed
* Traditional arguments against Aeschylus' authorship
    * Zeus depicted as a cruel tyrant
    * slightly different versification patterns
    * the third actor needed in one of the scenes




## stylometry: where it (usually) begins

```{r, echo = FALSE, message = FALSE}
library(stylo)
load("R_datasets/greek_drama_freqs.RData")
round(greek_drama[1:19,1:10], 3)
```




## table of computed distances

```{r, echo = FALSE, message = FALSE}
round(as.matrix(dist.delta(greek_drama)), 4)[1:19,1:5]
```



## hierarchical cluster analysis

```{r, echo = FALSE, message = FALSE, fig.align = "center"}
stylo(frequencies = greek_drama, gui = FALSE, titles.on.graphs = FALSE)
```



## multidimensional scaling

```{r, echo = FALSE, message = FALSE, fig.align = "center"}
stylo(frequencies = greek_drama, gui = FALSE, analysis.type = "MDS", titles.on.graphs = FALSE)
```



## bootstrap consensus tree

```{r, echo = FALSE, message = FALSE, fig.align = "center"}
stylo(frequencies = greek_drama, gui = FALSE, analysis.type = "BCT", mfw.min = 100, mfw.max = 1000, titles.on.graphs = FALSE)
```





# one variable vs. many



## simple logistic regression

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]

  node [shape = rectangle, style = filled, fontname = Helvetica]

  input_data [label = 'input\\ndata', fillcolor = Beige, width = 1]
  choice     [label = '🤔', fillcolor = LightYellow, width = 0.5]
  feature_1  [label = 'chosen\\nvariable', fillcolor = LightBlue, width = 0.5]
  classifier [label = 'classifier', fillcolor = LightYellow, width = 1, height = 2]
  yes_node   [label = 'yes', fillcolor = PaleGreen, width = 0.5]
  no_node    [label = 'no', fillcolor = MistyRose, width = 0.5]

  input_data -> choice
  choice     -> feature_1
  feature_1  -> classifier
  classifier -> yes_node
  classifier -> no_node

}")
```

## multivariate methods

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]

  node [shape = rectangle, style = filled, fontname = Helvetica]

  input_data [label = 'input\\ndata', fillcolor = Beige, width = 1]
  choice     [label = '🤔', fillcolor = LightYellow, width = 0.5]
  feature_1  [label = 'var 1', fillcolor = LightBlue, width = 0.5]
  feature_2  [label = 'var 2', fillcolor = LightBlue, width = 0.5]
  feature_3  [label = '. . .', fillcolor = LightBlue, width = 0.5]
  feature_n  [label = 'var n', fillcolor = LightBlue, width = 0.5]
  classifier [label = 'classifier', fillcolor = LightYellow, width = 1, height = 2]
  yes_node   [label = 'yes', fillcolor = PaleGreen, width = 0.5]
  no_node    [label = 'no', fillcolor = MistyRose, width = 0.5]

  input_data -> choice
  choice     -> feature_1
  choice     -> feature_2
  choice     -> feature_3
  choice     -> feature_n
  feature_1  -> classifier
  feature_2  -> classifier
  feature_3  -> classifier
  feature_n  -> classifier
  classifier -> yes_node
  classifier -> no_node

}")
```



## sentiment analysis

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]

  node [shape = rectangle, style = filled, fontname = Helvetica]

  input_data [label = 'input\\ntext', fillcolor = Beige, width = 1]
  choice     [label = '🤔', fillcolor = LightYellow, width = 0.5]
  feature_1  [label = 'charming', fillcolor = LightBlue, width = 0.5]
  feature_2  [label = 'awesome', fillcolor = LightBlue, width = 0.5]
  feature_3  [label = 'terrible', fillcolor = LightBlue, width = 0.5]
  feature_n  [label = '. . .', fillcolor = LightBlue, width = 0.5]
  classifier [label = 'classifier', fillcolor = LightYellow, width = 1, height = 2]
  yes_node   [label = 'positive', fillcolor = PaleGreen, width = 0.5]
  no_node    [label = 'negative', fillcolor = MistyRose, width = 0.5]

  input_data -> choice
  choice     -> feature_1
  choice     -> feature_2
  choice     -> feature_3
  choice     -> feature_n
  feature_1  -> classifier
  feature_2  -> classifier
  feature_3  -> classifier
  feature_n  -> classifier
  classifier -> yes_node
  classifier -> no_node

}")
```


## authorial recognition

```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]

  node [shape = rectangle, style = filled, fontname = Helvetica]

  input_data [label = 'input\\ndata', fillcolor = Beige, width = 1]
  choice     [label = '🤔', fillcolor = LightYellow, width = 0.5]
  feature_1  [label = 'the', fillcolor = LightBlue, width = 0.5]
  feature_2  [label = 'of', fillcolor = LightBlue, width = 0.5]
  feature_3  [label = 'in', fillcolor = LightBlue, width = 0.5]
  feature_n  [label = '. . .', fillcolor = LightBlue, width = 0.5]
  classifier [label = 'classifier', fillcolor = LightYellow, width = 1, height = 2]
  M_node   [label = 'Milton', fillcolor = MistyRose, width = 0.5]
  C_node    [label = 'Chaucer', fillcolor = PaleGreen, width = 0.5]
  S_node   [label = 'Shakespeare', fillcolor = MistyRose, width = 0.5]
  J_node   [label = 'Johnson', fillcolor = MistyRose, width = 0.5]

  input_data -> choice
  choice     -> feature_1
  choice     -> feature_2
  choice     -> feature_3
  choice     -> feature_n
  feature_1  -> classifier
  feature_2  -> classifier
  feature_3  -> classifier
  feature_n  -> classifier
  classifier -> M_node
  classifier -> C_node
  classifier -> S_node
  classifier -> J_node

}")
```



## structured data: literary texts


```{r, echo = FALSE, message = FALSE}
library(stylo)
data(galbraith)
data(lee)
round(galbraith[1:19,1:8], 3)
```

## structured data: iris

```{r}
iris
```

## classification methods

- the concept of multidimensionality & distances: [presentation](https://computationalstylistics.github.io/presentations/distances_intro/)
- crush intro to supervised classification (presentation in PDF)


#

![](DigiTS-RGB-short-SMALL.svg)


